---
title: "Smilator_omics"
author: "Anastasia&Nasimeh"
date: "6/27/2022"
output: html_document
---
# Simulator code for data generation process of omics data. In this case methylome. 


## SimMethyl

# Purpose:
# Simulating methylation data based on real-world data. Replicates the data generating process of real-world data using simulation parameters (e.g. number of samples) and generates a synthetic dataset with induced treatment effects.

# Input parameters
#1) beta matrix with sample as columns and CpGs as rows.
#2) values for the five parameters, where two out of five will be varied, while the rest will remain constant.

# Result:
# A beta matrix is equivalent to a real-world matrix with adjusted values.

```{r pressure, echo=FALSE}
library(schoolmath) # use of is.negative() 
library(GEOquery) #obtain the real-world data
library(zip) 
library(matrixStats) # calculate mean and SD based on dimension specified
```

```{r pressure, echo=FALSE}
# Function for calculating the mean of CpGs across all samples.
calculate_means <- function (beta_values_matrix) {
means <- rowMeans(beta_values_matrix, na.rm=TRUE)
means <- na.omit(means)
print(head(means))
return(means)
}
```

```{r pressure, echo=FALSE}
# Function for calculating the standard deviation of CpGs across all samples.
calculate_SD <- function (beta_values_matrix) {
SD <- matrixStats::rowSds(beta_values_matrix, na.rm=TRUE)
SD <- na.omit(SD)
print(head(SD))
return(SD)
}
```


```{r pressure, echo=FALSE}
# obtain indices for sampled CpGs to generate desirable number of CpGs in the simulated data.
set.seed(42)

user_specified_num_elements <- function(means, user_specified_n_CpGs){
sampling_indices = sample(1:dim(means)[1],user_specified_n_CpGs, replace = TRUE)
print(head(sampling_indices))
return(sampling_indices)
}


```


```{r pressure, echo=FALSE}
# Induce differences between groups by changing mean beta-values of selected CpGs. The modified CpGs are sampled from mean values of the reference group (control group). The output is the recorded indices of the truly modified CpGs and vector of means with introduced corresponding differences.

induce_group_differnces <- function(num_true_modified, vector_of_ref_means, effect_size, user_path){ 
set.seed(42)
  
truly_different_indices <- sample(1:length(vector_of_ref_means), size=num_true_modified, replace = FALSE)
filename <- paste0(user_path, '/truly_different_sites_indices.txt')
write.table(truly_different_indices, filename, row.names = FALSE, col.names = FALSE, sep = " ")
vector_of_affected_means <- vector_of_ref_means
vector_of_affected_means[truly_different_indices] <- vector_of_ref_means[truly_different_indices]+effect_size
vector_of_affected_means <- ifelse(vector_of_affected_means>1, 1, vector_of_affected_means) # keep modified means values within the range (0 and 1) specific to beta-value
return(list(vector_of_affected_means,truly_different_indices))
}
```



```{r pressure, echo=FALSE}
# get number of samples per group based on the control-case balance (healthy proportion)
get_group_number = function(healthy_proportion, total_num_samples){
g1_number_of_samples = healthy_proportion*total_num_samples
g2_number_of_samples = total_num_samples-g1_number_of_samples
g1_g2 = c(g1_number_of_samples ,g2_number_of_samples)
names(g1_g2) = c('Reference','Diseased')
return(g1_g2)
}
```



```{r pressure, echo=FALSE}
# Simulate beta values for every CpG across all samples. Due to the range of the real beta values (0 and 1), we "truncate"/adjust the values within the range. 
sample_distribution = function(means_vector, sds_vector, number_of_samples_in_group){
  all_cpgs <- c()
  cpg_name <- c()
for(i in 1:length(means_vector)){
cpg_i <- rnorm(number_of_samples_in_group, mean = means_vector[i], sd = sds_vector[i])
cpg_name[i] <- paste0("cpg",i)
all_cpgs <- rbind(all_cpgs,cpg_i)
#print(all_cpgs)
}
for (j in 1: length(all_cpgs)){
  if (all_cpgs[j] > 1){
  all_cpgs[j] = max(beta_matrix, na.rm = T)
  } else if (all_cpgs[j]< 0 || is.negative(all_cpgs[j])){
 all_cpgs[j]= min(beta_matrix, na.rm=T)
  } 
  }

rownames(all_cpgs) <- cpg_name
print(head(all_cpgs))
return(all_cpgs)
}
```


```{r pressure, echo=FALSE}
# Simulate methylation data for every group based on the shape parameters specific to the group (different means and common standard deviation). The output is data frame with all samples and all CpGs.
generate_cpgs_for_groups = function(g1_means_vector, g2_means_vector, vector_sd, g1_number_of_samples, g2_number_of_samples){
g1_cpgs <- sample_distribution(g1_means_vector, vector_sd, g1_number_of_samples)
g2_cpgs <- sample_distribution(g2_means_vector, vector_sd, g2_number_of_samples)
cpgs <- cbind(g1_cpgs, g2_cpgs)
print(head(cpgs))
return(cpgs)
}
```

```{r pressure, echo=FALSE}
# Generate group-specific sample names
generate_col_names = function(data_frame,g1_number_of_samples,g2_number_of_samples){
  g1_sample_names= paste0("Gr1_Samp",1:g1_number_of_samples)
  g2_sample_names= paste0("Gr2_Samp",1:g2_number_of_samples)
  all_sample_names = c(g1_sample_names, g2_sample_names)
}
```

```{r pressure, echo=FALSE}
# Create all combinations (cartisian product) of the input parameters (5 user-defined parameters)
get_all_combinations = function(total_num_samples_vector,effect_size_vector,healthy_proportion,num_true_modified,user_specified_n_CpGs,user_path, zip_file_name){
all_combinations = expand.grid(total_num_samples_vector,effect_size_vector,healthy_proportion,num_true_modified,user_specified_n_CpGs)
all_combinations$Va6 = user_path
all_combinations$Var7 = paste0(zip_file_name,1:nrow(all_combinations))
colnames(all_combinations) = c("n_samples","effect_size","healthy_proportion","n_true_modified_CpGS","n_CpGs","user_path","zip_file_order")
write.csv(all_combinations,"all_combinations.csv", row.names = FALSE)
print(head(all_combinations))
return(all_combinations)
}

```

```{r pressure, echo=FALSE}
# Save data as a zip-file with standardized names. A zip file is equal to one simulation data based on a specific set of parameters (specific combination). A zip file contains 3 files: "Simulated_data", "truly_different_sites_indices", "User_Parameters". 
create_zip = function(user_path,zip_file_name){
  path_directory = user_path
  zipfile = paste0(user_path,"/",zip_file_name,".zip")
  my_files = c(paste0(user_path,"/Simulated_data.txt"),paste0(user_path,"/truly_different_sites_indices.txt"),paste0(user_path,"/User_Parameters.txt"))
  zip::zipr(zipfile = zipfile, files = my_files) 
}
```



```{r pressure, echo=FALSE}
# Aggregated (main) function which generates simulated data set for one environmental set-up (user-defined). The input is a vector with values for every parameter defining the environmental/experimental set-up. The output is the zip file with simulated data, indices of the truly modified CpGs, and input parameters of the user. 
simulator = function(combined_param_vector){ 
total_num_samples = as.numeric(combined_param_vector[1])
effect_size = as.numeric(combined_param_vector[2])
healthy_proportion = as.numeric(combined_param_vector[3])
num_true_modified = as.numeric(combined_param_vector[4])
user_specified_n_CpGs = as.numeric(combined_param_vector[5])
user_path = combined_param_vector[6]
zip_file_name = combined_param_vector[7]
means_real_world = calculate_means(beta_matrix)
SD_real_world = calculate_SD(beta_matrix)
means_SD_real_world = cbind(means_real_world,SD_real_world)


indices = user_specified_num_elements(means_SD_real_world, user_specified_n_CpGs)
means_SD = means_SD_real_world[indices,] 
vector_of_ref_means = means_SD[,1]
vector_sd = means_SD[,2]


list_indices_affected_means = induce_group_differnces(num_true_modified,vector_of_ref_means,effect_size,user_path)
vector_of_affected_means = list_indices_affected_means[[1]]

g1_number_of_samples=get_group_number(healthy_proportion, total_num_samples)[1]
g2_number_of_samples=get_group_number(healthy_proportion, total_num_samples)[2]


simulated_data = generate_cpgs_for_groups(vector_of_ref_means,vector_of_affected_means, vector_sd, g1_number_of_samples, g2_number_of_samples)

colnames(simulated_data) = generate_col_names(simulated_data,g1_number_of_samples,g2_number_of_samples) #check this 

file_name_simulated_data = paste0(user_path,"/Simulated_data.txt")
write.table(simulated_data, file_name_simulated_data,col.names = TRUE, row.names = TRUE)

param_names = c("Total number of samples","User-spesified number of CpGs","Healthy proportion","Effect size","Number of true modified CpG sites")
param_values = c(total_num_samples,user_specified_n_CpGs,healthy_proportion,effect_size,num_true_modified)
param_summary <- as.data.frame(matrix(nrow = length(param_names), ncol = 2))

names(param_summary) = c("Parameter","Value")
param_summary$Parameter = param_names
param_summary$Value = param_values

file_name_parameters = paste0(user_path,"/User_Parameters.txt")
write.table(param_summary,file_name_parameters,row.names = FALSE, col.names = TRUE, sep = " ")

create_zip(user_path,zip_file_name)
}

```


```{r pressure, echo=FALSE}
# Function calls simulator to generate multiple simulations by alterating through the data frame containing parameter values for all possible combinations. Every row in the data frame represents a specific scenario/environmental set-up.
multi_SimMethyl = function(total_num_samples_vector,effect_size_vector,healthy_proportion,num_true_modified,user_specified_n_CpGs,user_path,zip_file_name){
combination_df = get_all_combinations(total_num_samples_vector,effect_size_vector,healthy_proportion,num_true_modified,user_specified_n_CpGs,user_path,zip_file_name)
final_output <- apply(combination_df, 1, simulator)
return(final_output)
}

```

```{r pressure, echo=FALSE}
#Choose the real-world data 
# In this case we have used a data set with two groups (healthy versus patients with tuberculosis). Beta matrix data include Immumina array with 850 000 cpG probes and 32 samples. 
#Source: https://www.jci.org/articles/view/134622#SEC4
#Tissue type: PBMCs
 
GEO_real_world_data <- "GSE145714" # ID for real world data 

gset <- getGEO(GEO_real_world_data, GSEMatrix =TRUE, getGPL=FALSE, destdir="example_dataset/")

gset <- gset$GSE145714_series_matrix.txt.gz

clinical_data <- pData(gset) # clinico-pathological description

clinical_data <- clinical_data[,c(2,41:44)]

colnames(clinical_data)[c(2:5)] <- c("hiv_status", "gender", "tb_status", "timepoint")

clinical_data <- clinical_data[order(clinical_data$tb_status),]

beta_matrix <- exprs(gset)

#Creating a subset of the beta_matrix (to reduce running time during testing)
beta_matrix <- beta_matrix[sample(1:nrow(beta_matrix),1000,replace = FALSE),sample(1:ncol(beta_matrix),10,replace = FALSE)]

rm(gset)
```

```{r pressure, echo=FALSE}
start.time <- Sys.time()
print(paste0("Start Time ", start.time))

user_path = getwd() #"/Users/sima/Documents/summer project/Simulator")
zip_file_name ="SimMethyl_run_"

# User-defined parameter input. Specify testing range as vector. The data will be saved in the same directory as working directory. 
healthy_proportion <- 0.5
num_true_modified <- 30

user_specified_n_CpGs <- 500

total_num_samples_vector <- c(20,60,100)
effect_size_vector <- c(0.02,0.04,0.06)

multi_SimMethyl(total_num_samples_vector,effect_size_vector,healthy_proportion,num_true_modified,user_specified_n_CpGs,user_path,zip_file_name)

end.time <- Sys.time()
print(paste0("End Time ", end.time))
time.taken <- end.time - start.time
print(paste0("Time Taken ", time.taken))
```


